---
title: "Interpreting an ML model with SHAP values"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/bioinformatics_intro_workshop

## Introduction

In this workshop, we will start working on some RNA-seq data. Here is the citation:

Cho JW, Shim HS, Lee CY, Park SY et al. The importance of enhancer methylation for epigenetic regulation of tumorigenesis in squamous lung cancer. Exp Mol Med 2022 Jan;54(1):12-22. PMID: 34987166

```{r setup}

suppressPackageStartupMessages({
  library("kableExtra")
  library("DESeq2")
  library("gplots")
  library("beeswarm")
  library("parallel")
  library("kableExtra")
  library("xgboost")
  library("iml")
  library("ggplot2")
  library("beeswarm")
})

```

## Download gene expression counts and read the file

Download from NCBI.

```{r,dl}

DATA="GSE158420.counts.txt.gz"

if ( ! file.exists(DATA) ) {
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE158420&format=file&file=GSE158420%5Fcounts%2Etxt%2Egz",
    destfile=DATA)
}

```

Read data.
Note that there are a couple of duplicated gene symbols.
Here I use `aggregate()` command to deal with them.

```{r,read_data}

mx <- read.table(DATA,row.names = NULL)
str(mx)
which(duplicated(mx$row.names))
xa <- aggregate(. ~ row.names,mx,sum)
rownames(xa) <- xa$row.names
str(xa)
xa$row.names = NULL

head(xa)

```

## Basic QC - number of reads per sample

Less than 10 million is not acceptable. More than 20 million is considered good.

```{r,sample_count_qc,fig.width=5,fig.height=5}

mycolsums <- colSums(xa)

mycolsums
summary(mycolsums)
mycolsums <- sort(mycolsums)
head(mycolsums,10)
tail(mycolsums,10)
barplot(mycolsums)


barplot(head(mycolsums))

```

# Does this dataset need filtering of lowly expressed genes?

There are ~60k genes in the main annotation sets.
We can count the number of rows to see how many genes included in this dataset.

I also run my own threshold, which is average of 10 reads per sample.

```{r,lowgenes}

dim(xa)
hist(rowMeans(xa))
table(rowMeans(xa)>=10)

xf <- xa[which(rowMeans(xa)>=10),]
dim(xa)
dim(xf)

```

# Principal component analysis

Use PCA to understand the overall variation between samples in the dataset.

```{r,pca1,fig.height=7}

mds <- cmdscale(dist(t(scale(xf))),k=10)
barplot(colSums(abs(mds)),names.arg = 1:10)

head(mds)

plot(mds[,1:2],pch=19,col="gray",
     main="Dim 1 and 2",xlab="PCA1",ylab="PCA2")
text(mds[,1],mds[,2],labels = colnames(xf))

plot(mds[,c(1,3)],pch=19,col="gray",
      main="Dim 1 and 3",xlab="PCA1",ylab="PCA3")
text(mds[,1],mds[,3],labels = colnames(xf))

plot(mds[,2:3],pch=19,col="gray",
      main="Dim 2 and 3",xlab="PCA2",ylab="PCA3")
text(mds[,2],mds[,3],labels = colnames(xf))

```

# Curate samplesheet

The samplesheet is the sample level metadata, which we need for the differential expression step.

```{r,ss}

ss <- as.data.frame(colnames(xa))
head(ss)

ss$tumor <- as.numeric(grepl("T",ss[,1]))

ss$patient <- gsub("T","" , gsub("N","",ss[,1]))

rownames(ss) <- ss[,1]

head(ss)

```

## DE

We will use DESeq2 for differential expression.

```{r,de1}

dds <- DESeqDataSetFromMatrix(countData = xf , colData=ss, design = ~ patient + tumor)
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])

head(dge) |> kbl(caption="Top significant genes") |> kable_paper("hover", full_width = F)

```

## Heatmap visualisation

Heatmaps are great for showing high dimensional data.

```{r,hm1}

topgenes <- head(rownames(dge),50)

colfunc <- colorRampPalette(c("blue", "white", "red"))

rpm <- apply(xf,2,function(x) {x/sum(x) * 1e6 } )

top <- rpm[which(rownames(rpm) %in% topgenes),]

csc = as.character(ss$tumor)

heatmap.2(as.matrix(top),trace="none",col=colfunc(25),scale="row",
    margins = c(6,6), cexRow=0.6, cexCol=0.3, ColSideColors=csc )

```

You may want to look at a few genes in more detail.

```{r,bs1}

myrows <- c(1,2,3,5,6,38)

par(mfrow=c(2,3))

nullplot <- lapply(myrows,function(i) {
  nam=rownames(top)[i]
  gdata <- top[i,]
  ndat <- gdata[grep("N",names(gdata))]
  tdat <- gdata[grep("T",names(gdata))]
  plotdat <- list("non-tumor"=ndat,"tumor"=tdat)
  boxplot(plotdat,ylab="normalised expression (RPM)",cex=0,col="white",main=nam)
    beeswarm(plotdat,add=TRUE,cex=0.9,pch=19)
})

nullplot <- lapply(myrows,function(i) {
  nam=rownames(top)[i]
  gdata <- log10(top[i,]+0.1)
  ndat <- gdata[grep("N",names(gdata))]
  tdat <- gdata[grep("T",names(gdata))]
  plotdat <- list("non-tumor"=ndat,"tumor"=tdat)
  boxplot(plotdat,ylab="log10 (RPM)",cex=0,col="white",main=nam)
    beeswarm(plotdat,add=TRUE,cex=0.9,pch=19)
})

par(mfrow=c(1,1))

```

## Prep data for ML

Using the top 50 DESEQ2 genes as input to train xgboost model.

I will reserve 20 patients for the test.
The rest will be used in the training.

```{r,mlprep1}

head(top)

my_data <- as.data.frame(t(top))

my_labels <- as.numeric(grepl("T",rownames(my_data)))

Ntest=20
Ntrain=nrow(my_data) - Ntest

test_data <- my_data[1:Ntest,]
test_labels <- my_labels[1:Ntest]

train_data <- my_data[(Ntest+1):nrow(my_data),]
train_labels <- my_labels[(Ntest+1):length(my_labels)]

dim(test_data)
dim(train_data)
length(test_labels)
length(train_labels)

```

## Train model

```{r,train1}

model <- xgboost(x = train_data, y = train_labels, nrounds = 100, objective = "reg:squarederror")

model

```

## Predict

```{r,pred1}

pred <- predict(model, test_data)
prediction <- as.numeric(pred > 0.5)
table(test_labels == prediction)
table(test_labels == prediction)[2] / length(prediction)

```

## Calculate shap values

`iml` package is used to calculate SHAP values.

```{r,shap1}

# Create a Predictor object
predictor <-  iml::Predictor$new(model, data = train_data)

# Example for how to calculate SHAP values for one observations
shapley <- iml::Shapley$new(predictor, x.interest = train_data[1, ])
shapley

# Now calculate all SHAP values
# mclapply causes this to freeze, so we need to do it single threaded for now
shapley <- lapply(1:nrow(train_data), function(i) {
  iml::Shapley$new(predictor, x.interest = train_data[i, ])
})

shapley[[1]]

shap_list <- lapply(1:length(shapley), function(i) { shapley[[i]]$results })
shap_df <- do.call(rbind, shap_list)
meanabs <- function(x) { mean(abs(x)) }
shap_dfa <- aggregate(phi ~ feature,shap_df,FUN=meanabs )
shap_dfa <- shap_dfa[order(-shap_dfa$phi),]
head(shap_dfa)

```

In this simple model, only the SERPINE2 gene was used to classify the tumor and control samples. So it is not a great example.

## Remove SERPINE2 from data

Let's see if the model is different or performs worse than the original one.

```{r,remove_serpine2}

train_data2 <- train_data[,-which(colnames(train_data)=="SERPINE2")]
model2 <- xgboost(x = train_data2, y = train_labels, nrounds = 100, objective = "reg:squarederror")
test_data2 <- test_data[,which(colnames(test_data)!="SERPINE2")]

pred2 <- predict(model2, test_data2)
prediction2 <- as.numeric(pred2 > 0.5)
table(test_labels == prediction2)
table(test_labels == prediction2)[2] / length(prediction2)

predictor2 <-  iml::Predictor$new(model2, data = train_data2)
shapley2 <- iml::Shapley$new(predictor2, x.interest = train_data2[1, ])

shapley2 <- lapply(1:nrow(train_data2), function(i) {
  iml::Shapley$new(predictor2, x.interest = train_data2[i, ])
})

shapley2[[1]]

shap_list2 <- lapply(1:length(shapley2), function(i) { shapley2[[i]]$results })
shap_df2 <- do.call(rbind, shap_list2)
shap_dfa2 <- aggregate(phi ~ feature,shap_df2,FUN=meanabs )
shap_dfa2 <- shap_dfa2[order(-shap_dfa2$phi),]
head(shap_dfa2)

```

Barplot.

```{r,shap_barplot}

modelgenes <- head(shap_dfa2,10)$feature
modelweights <- head(shap_dfa2,10)$phi
names(modelweights) <- modelgenes

par(mar=c(5.1,8.1,4.1,2.1))
barplot(rev(modelweights),horiz=TRUE,las=1,xlab="Mean Abs Phi values",main="Model weights")
par(mar=c(5.1,4.1,4.1,2.1))

```

## Interpret SHAP values

The Shapley object contains the contribution of each feature to the prediction for the specific observation.

each feature has a corresponding SHAP value, indicating its contribution to the model's prediction
for that observation.
Positive values indicate that the feature increased the prediction, while negative values suggest it
decreased the prediction.

```{r,shap2}

print(shapley$results) |>
    kbl(caption = "Shapley values") |>
    kable_paper("hover", full_width = F)

```

Now create a summary plot

```{r,shap_plot1}

shapley$plot()

```

## Remove SERPINE2 from data

```{r,serpine1_remove}

train_data2 <- train_data[,-which(colnames(train_data)=="SERPINE2")]
model2 <- xgboost(x = train_data2, y = train_labels, nrounds = 100, objective = "reg:squarederror")
pred2 <- predict(model2, test_data)
prediction2 <- as.numeric(pred2 > 0.5)
table(test_labels == prediction2)
table(test_labels == prediction2)[2] / length(prediction2)

predictor2 <-  iml::Predictor$new(model2, data = train_data2)
shapley <- iml::Shapley$new(predictor2, x.interest = train_data2[1, ])

```

Now look at the direction of the SHAP values as well.

SHAP values for normal and cancer samples are separated for charting.

```{r,interpret2}

shap_l <- lapply(topgenes,function(x) { shap_df2[which(shap_df2$feature == x),"phi"] } )
names(shap_l) <- topgenes

which(lapply(shap_l,meanabs)>0)

mytopmodelgenes <- names(which(lapply(shap_l,meanabs)>0))
shap_top <- shap_l[which(names(shap_l) %in% mytopmodelgenes)]
shap_top <- shap_top[order(unlist(lapply(shap_top,median)))]

shap_top

boxplot(shap_top,cex=0,horizontal=TRUE,las=1,col="white",main="Phi values for training data")
beeswarm(shap_top,horizontal=TRUE,pch=19,cex=0.4,add=TRUE,col="darkgray")

LEN=54
seq(1,LEN,2)

NORM <- lapply(shap_top,function(x) { x[seq(1,LEN,2)] })
CANC <- lapply(shap_top,function(x) { x[seq(1,LEN,2)+1] })
boxplot(shap_top,cex=0,horizontal=TRUE,las=1,col="white",main="Phi values for training data")
mtext("blue=normal, red=cancer sample")
beeswarm(NORM,horizontal=TRUE,pch=19,cex=0.4,add=TRUE,col="blue")
beeswarm(CANC,horizontal=TRUE,pch=19,cex=0.4,add=TRUE,col="red")

```

From this we can see that of all the genes in this model, RGCC is the most inflential. Higher expression of RGCC is associated with cancer classification. RGCC, LMO7, CLIC5 and CALML3 are associated with cancer, while higher AGER is associated with normal tissue.

## Conclusion

Here, we trained an XGboost model on a small number of lung cancer datasets (54) using only the top 50 genes from a limma analysis.
Testing on a separate 20 samples found it to be approximately 95% accurate.
We used SHAP analysis to understand how strongly the genes influence the model.
Initially the model exclusively used SERPINE2 to classify.
Another model was generated after removing SERPINE2 and it showed RGCC, CALML3, LMO7, CLIC5, AGER and ZIC2 could also be informative in a model.
This second model was also 95% accurate.
This small example shows how a panel of genes can be selected from an ML model.
The SHAP analysis here can be applied to different ML models such as random forest, support vector machine, etc.
This is one method to interpret ML models.
This type of approach could be used for various omics data such as proteomics, microbiome, or DNA methylation.

## Session information

```{r,session}

save.image("lung_cancer_gene_expression_shap.Rdata")

sessionInfo()

```
